The effects of host galaxy properties on merging compact binaries detectable by LIGO

### 2016-07-12 paper/paper.tex
UPDATE updates

compact binary populations.  As an example, we demonstrate how the coalescing binary black hole population can differ by

@@

\section{Introduction}
% POINT: LIGO soon will detect compact binary mergers....
%     ... and possibly EM counter part too
%     ... which will enable new physics
Gravitational waves have been detected from  coalescing  black hole binaries \citep{DiscoveryPaper,LIGO-O1-BBH}. 
Over the next few years, ground-based gravitational wave detectors like LIGO 
and Virgo  should detect the gravitational
wave signal from many more similar merging compact binaries \citep{LIGO-O1-BBH,RatesPaper,gwastro-EventPopsynPaper-2016}, as well as  binary  neutron stars and black hole-neutron star binaries \citep{LIGO-Inspiral-Rates,popsyn-LowMetallicityImpact2c-StarTrackRevised-2014}.
%
The host galaxies of gravitational wave sources will be identified, either directly or statistically.  
On the one hand, addition to the strong gravitational wave signal, mergers of neutron stars are expected to  frequently be accompanied by detectable
electromagnetic radiation  via a number of  mechanisms, including a strong but tightly beamed
ultrarelativistic jet and weakly-beamed but more isotropic thermal radiation from an expanding hot shell
\citep[see,e.g.,][and references therein]{2013PhRvL.111r1101C,short-grb-GWCoincidenceEM-MetzgerBerger2011}.  
% People we can't afford the space to cite, alas:
%    - 2015arXiv150101986H   -
%    - 2013MNRAS.430.2121P - Rosswog recent
%    - Shibata group/Koutarou
In addition to providing a wealth of information about the merger event itself, a multimessenger detection will pin down
the sky position and therefore approximate birthplace of each merging binary.    
%
On the other hand, gravitational wave detectors can perform high-precision sky and volumetric localization \citep{2016LRR....19....1A,2016arXiv160307333S}.  As GW detector
networks increase in number and sensitivity, these localizations will allow statistical and, eventually, unique identification of
host galaxies directly, even without associated electromagnetic emission.
%
As with supernovae and  GRBs, these host galaxy associations   are expected to tightly constrain models for compact binary
formation; see, for comparison,  \citet{2011MNRAS.412.1508M}, \citet{long-grb-GuettaPiran2007},
\citet{2014ARAA..52...43B}, and references therein. %2010ApJ...722.1879M,2010MNRAS.407.1314M}.     
%

@@

This paper is organized as follows. 
In \S \ref{sec:sims}, we describe detailed hydrodynamical simulations of  several galaxies, including two of Milky Way-mass and two dwarfs.  Though the two Milky Way-like galaxies are morphologically similar at $z = 0$, we
show that their star formation and chemical evolution history have subtle but important differences due to their
distinctive merger histories.   To demonstrate
these differences have a significant practical impact,  in \S \ref{sec:model}, we introduce a simple,
metallicity-dependent  phenomenological model to calculate the present-day rate and mass distribution of compact binary
mergers from  a galaxy's known history.   In \S \ref{sec:results:BBH}, we use this model to demonstrate the binary black
hole (BH-BH) coalescence rate
can depend sensitively on the galaxy's assembly history, in a manner which is far from apparent from its present-day
properties.    We also demonstrate that even the coalescence rates of binary neutron stars (NSNS) and black hole-neutron
star (BH-NS) binaries also depend surprisingly sensitively on the assembly history of each galaxy. 
In \S~\ref{sec:Discussion} we discuss the implications of our study for the interpretation of host galaxy associations
identified via transient multmiessenger astronomy, in the short and long term.  We summarize our results in \S~ \ref{sec:conclude}.
%% In \S \ref{sec:results} we show that while these two galaxy formation models should produce comparable detectable binary neutron star
%% populations, these two scenarios could produce dramatically different populations of electromagnetically detectable
%% merging black hole-neutron star binaries. 
%

@@

%\editremark{recheck numbers: we are tweaking things, and Zsun is seemingly inconsistent in the two approaches }
 Our two galaxies have extremely similar star formation histories and metallicity evolution, particularly at late
 times.  
However, at early times, the \ExcitingGalaxy{} galaxy forms stars for $\simeq 2\unit{Gyr}$ at a lower characteristic metallicity
($Z\simeq 0.8 \times10^{-3}$) compared
to the \BoringGalaxy{} galaxy ($Z\simeq 2\times 10^{-3}$).  In this reigme, the formation efficiency $\lambda$ and sensitive volume $V$
can depend sensitively on mass; for example, for binary black holes, the ratio of  $(\lambda V)_{\BoringGalaxy{}}/(\lambda
V)_{\ExcitingGalaxy{}} \simeq 7$.    However, in this same epoch, the star formation rate in the \ExcitingGalaxy{} (low-metallicity)
galaxy is smaller, by a factor of roughly 2.
% 1e-3:  V = 276, lambda= 400   
% 2e-3:  V= 152.657625157, lambda = 100
% so the ratio should be a factor of about 7x higher due to this region.
Therefore, because only a fraction of order $10\%$ \textbf{[checkme-JILLIAN?]} of all star formation occurs in this epoch, by this order of magnitude
argument, the overall number of
present-day coalescing binary black holes in our two galaxies is expected to differ by of order ten percent.
% 0.9 + 0.1*7/2
%
Using the concrete phenomenological calculations described above, we in fact find
$(r_{D}/M)_{\BoringGalaxy{}}/(r_{D}/M)_{\ExcitingGalaxy{}}\simeq 0.9$.

The close agreement between the two galaxies' binary black hole populations  is entirely a
a numerical coincdence between the specific star formation rate and metallicities in these two galaxies.  The effects described here
  are well within the scatter around the mass-metallicity relation.  Marginally different realizations of these
  histories can easily produce factors of order unity difference in galaxies with otherwise indistinguishable
  present-day properties.\editremark{do we have another example}

@@

For binary black hole mergers in particular, we show that the present-day binary black hole coalescence rate can vary by factors of several in otherwise
similar present-day environments.   Our calculations support and extend prior investigations which demonstrated both
that black hole merger rates depend sensitively on low-metallicitiy environments
\citep[see,e.g.][]{PSellipticals,popsyn-LowMetallicityImpact2b-StarTrackRevised-2013,gwastro-EventPopsynPaper-2016}.   More broadly,
because compact binaries can merge long after they form, their host galaxy can evolve substantially in composition
between birth and merger.
%% and that the delay between compact
%% binary formation and coalesence ccn be
%% the sensitive dependence of binary evolution on
%% metallicity implies that
%% * BBH properties far from apparent.  No surprise: delays can be long , particularly because formation in low Z favored,
%% so system changes significantly from assembly conditions

>>>>

compact binary populations \jillianremark{ UPDATE}.  As an example, we demonstrate how the coalescing binary black hole population can differ by

@@

\section{Introduction}
% POINT: LIGO soon will detect compact binary mergers....
%     ... and possibly EM counter part too
%     ... which will enable new physics
Gravitational waves have been detected from  coalescing  black hole binaries \citep{DiscoveryPaper,LIGO-O1-BBH}. 
Over the next few years, ground-based gravitational wave detectors like LIGO 
and Virgo  should detect the gravitational
wave signal from many more similar merging compact binaries \citep{LIGO-O1-BBH,RatesPaper,gwastro-EventPopsynPaper-2016}, as well as  binary  neutron stars and black hole-neutron star binaries \citep{LIGO-Inspiral-Rates,popsyn-LowMetallicityImpact2c-StarTrackRevised-2014}.
%
The host galaxies of gravitational wave sources will be identified, either directly or statistically.  
If the event is a merger of at least one neutron star, it is expected to be accompanied by detectable
electromagnetic radiation  via a number of  mechanisms \citep[see][and references therein]{Metzger16} in addition to the strong gravitational wave signal. % including a strong but tightly beamed
%ultrarelativistic jet and weakly-beamed but more isotropic thermal radiation from an expanding hot shell
%\citep[see,e.g.,][and references therein]{2013PhRvL.111r1101C,short-grb-GWCoincidenceEM-MetzgerBerger2011}.  
% People we can't afford the space to cite, alas:
%    - 2015arXiv150101986H   -
%    - 2013MNRAS.430.2121P - Rosswog recent
%    - Shibata group/Koutarou
%  In addition to providing a wealth of information about the merger event itself, 
A multimessenger detection will pin down
the sky position and therefore approximate birthplace of each merging binary \cite{Nissanke13}.    
%
Even in the absence of well-identified electromagnetic counterparts, host galaxy information is still available from gravitational wave localization alone \citep{2016LRR....19....1A,2016arXiv160307333S}.  As GW detector
networks increase in number and sensitivity, these localizations will allow statistical and, eventually, unique identification of
host galaxies directly, even without associated electromagnetic emission.
%
As with supernovae and  GRBs, these host galaxy associations   are expected to tightly constrain models for compact binary
formation; see, for comparison,  \citet{2011MNRAS.412.1508M}, \citet{long-grb-GuettaPiran2007},
\citet{2014ARAA..52...43B}, and references therein. %2010ApJ...722.1879M,2010MNRAS.407.1314M}.     
%

@@

This paper is organized as follows. 
In \S \ref{sec:sims}, we describe detailed hydrodynamical simulations of  several galaxies, including two of Milky Way-mass and two dwarfs.  Though the two Milky Way-like galaxies are morphologically similar at $z = 0$, we
show that their star formation and chemical evolution history have subtle but important differences due to their
distinctive merger histories.   To demonstrate
these differences have a significant practical impact,  in \S \ref{sec:model}, we introduce a simple,
metallicity-dependent  phenomenological model to calculate the present-day rate and mass distribution of compact binary
mergers from  a galaxy's known history.   In \S \ref{sec:results:BBH}, we use this model to demonstrate the binary black
hole (BH-BH) coalescence rate
can depend sensitively on the galaxy's assembly history, in a manner which is far from apparent from its present-day
properties. \jillianremark{ UPDATE}   We also demonstrate that even the coalescence rates of binary neutron stars (NSNS) and black hole-neutron
star (BH-NS) binaries also depend surprisingly sensitively on the assembly history of each galaxy. 
In \S~\ref{sec:Discussion} we discuss the implications of our study for the interpretation of host galaxy associations
identified via transient multmiessenger astronomy, in the short and long term.  We summarize our results in \S~ \ref{sec:conclude}.
%% In \S \ref{sec:results} we show that while these two galaxy formation models should produce comparable detectable binary neutron star
%% populations, these two scenarios could produce dramatically different populations of electromagnetically detectable
%% merging black hole-neutron star binaries. 
%

@@

%\editremark{recheck numbers: we are tweaking things, and Zsun is seemingly inconsistent in the two approaches }
 Our two galaxies have extremely similar star formation histories and metallicity evolution, particularly at late
 times.  
However, at early times, the \ExcitingGalaxy{} galaxy forms stars for $\simeq 2\unit{Gyr}$ at a lower characteristic metallicity
($Z\simeq 0.8 \times10^{-3}$) compared
to the \BoringGalaxy{} galaxy ($Z\simeq 2\times 10^{-3}$).  In this reigme, the formation efficiency $\lambda$ and sensitive volume $V$
can depend sensitively on mass; for example, for binary black holes, the ratio of  $(\lambda V)_{\BoringGalaxy{}}/(\lambda
V)_{\ExcitingGalaxy{}} \simeq 7$.    \jillianremark{ UPDATE} However, in this same epoch, the star formation rate in the \ExcitingGalaxy{} (low-metallicity)
galaxy is smaller, by a factor of roughly 2.
% 1e-3:  V = 276, lambda= 400   
% 2e-3:  V= 152.657625157, lambda = 100
% so the ratio should be a factor of about 7x higher due to this region.
Therefore, because only a fraction of order $10\%$ \textbf{[checkme-JILLIAN?]} of all star formation occurs in this epoch, by this order of magnitude
argument, the overall number of
present-day coalescing binary black holes in our two galaxies is expected to differ by of order ten percent.
% 0.9 + 0.1*7/2
%
Using the concrete phenomenological calculations described above, we in fact find
$(r_{D}/M)_{\BoringGalaxy{}}/(r_{D}/M)_{\ExcitingGalaxy{}}\simeq 0.9$.

The close agreement between the two galaxies' binary black hole populations \jillianremark{ UPDATE} is entirely a
a numerical coincdence between the specific star formation rate and metallicities in these two galaxies.  The effects described here
  are well within the scatter around the mass-metallicity relation.  Marginally different realizations of these
  histories can easily produce factors of order unity difference in galaxies with otherwise indistinguishable
  present-day properties.\editremark{do we have another example}

@@

For binary black hole mergers in particular, we show that the present-day binary black hole coalescence rate can vary by factors of several \jillianremark{ UPDATE} in otherwise
similar present-day environments.   Our calculations support and extend prior investigations which demonstrated both
that black hole merger rates depend sensitively on low-metallicitiy environments
\citep[see,e.g.][]{PSellipticals,popsyn-LowMetallicityImpact2b-StarTrackRevised-2013,gwastro-EventPopsynPaper-2016}.   More broadly,
because compact binaries can merge long after they form, their host galaxy can evolve substantially in composition
between birth and merger.
%% and that the delay between compact
%% binary formation and coalesence ccn be
%% the sensitive dependence of binary evolution on
%% metallicity implies that
%% * BBH properties far from apparent.  No surprise: delays can be long , particularly because formation in low Z favored,
%% so system changes significantly from assembly conditions
